---
title: "Bootstrap Profile Likelihood Confidence Intervals"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Bootstrap Profile Likelihood Confidence Intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svg"
)
```

```{r setup}
library(likelyprofiler)
set.seed(123)  # For reproducible results
```

## Bootstrap Threshold Method

The bootstrap method provides data-driven thresholds for profile likelihood confidence intervals instead of relying on asymptotic chi-squared approximations. This is particularly useful when the asymptotic assumptions may not hold.

## Example: Normal Distribution

We'll use a simple normal distribution example where both methods should give similar results, demonstrating the bootstrap implementation.

```{r}
# Generate synthetic normal data
n_data = 100
true_mu = 2.5
true_sigma = 1.2
data_normal = rnorm(n_data, true_mu, true_sigma)

# Set up bounds
bounds = list(lower = c(-1, 0.1), upper = c(6, 4))

# Define negative log-likelihood for normal distribution
negLogLikNormal = function(params, dataset = data_normal) {
  mu = params[1]
  sigma = params[2]
  if (sigma <= 0) return(1e10)  # Constraint: sigma > 0
  -sum(dnorm(dataset, mu, sigma, log = TRUE))
}

# Find ML estimates
ml_result = optim(c(mean(data_normal), sd(data_normal)), negLogLikNormal, 
                  dataset = data_normal, method = "L-BFGS-B", 
                  lower = bounds$lower, upper = bounds$upper)
ml_params = ml_result$par
names(ml_params) = c("mu", "sigma")
```

## Create Bootstrap Data Generator

Define a function to generate bootstrap datasets from the fitted distribution:

```{r}
# Create bootstrap data generation function
generateData = function(params) {
  # Generate new dataset from current parameter estimates
  rnorm(n_data, params[1], params[2])
}
```

## Compare Fixed vs Bootstrap Methods

```{r param_parallel_hidden, include=FALSE}
# Create a cluster for parameter-level parallelization (vignette building)
lib_likelyprofiler = dirname(path.package("likelyprofiler"))
cluster = parallel::makeCluster(2)
parallel::clusterExport(cluster, c("lib_likelyprofiler", "n_data"))
invisible(parallel::clusterEvalQ(cluster, library(likelyprofiler, lib.loc = lib_likelyprofiler)))
```

```{r param_parallel_user, eval=FALSE}
# Create a cluster for parameter-level parallelization
cluster = parallel::makeCluster(2)
parallel::clusterExport(cluster, "n_data")
```

```{r}
# Fixed threshold method (traditional)
result_fixed = computeLikelihoodProfiles(
  params_current = ml_params,
  negLogLikelihood = negLogLikNormal,
  bounds = bounds,
  profile_options = list(
    grid_points = 50,
    threshold_method = "fixed",
    ll_ratio_threshold = qchisq(0.95, 1)  # 95% CI
  ),
  cluster = cluster,
  dataset = data_normal,
  verbose = FALSE
)

# Bootstrap threshold method
result_bootstrap = computeLikelihoodProfiles(
  params_current = ml_params,
  negLogLikelihood = negLogLikNormal,
  bounds = bounds,
  profile_options = list(
    grid_points = 50,
    threshold_method = "bootstrap",
    bootstrap_conf_level = 0.95,
    n_bootstrap = 200
  ),
  generateData = generateData,
  cluster = cluster,
  dataset = data_normal,
  verbose = FALSE
)

parallel::stopCluster(cluster)
```

## Plot results

Plot bootstrap LRT statistics:

```{r}
# result_bootstrap$bootstrap_stats has dim (n_bootstrap, n_params, n_grid)
if (!is.null(result_bootstrap$bootstrap_stats)) {
  # Plot histogram of LRT statistics for first parameter at grid midpoint
  midpoint_idx = ceiling(dim(result_bootstrap$bootstrap_stats)[3] / 2)
  bootstrap_stats_to_plot = result_bootstrap$bootstrap_stats[, 1, midpoint_idx]
  q95_chisquared = qchisq(0.95, 1)
  q95_observed = quantile(bootstrap_stats_to_plot, probs = 0.95)
  hist(bootstrap_stats_to_plot, 
       main = "Bootstrap LRT Statistics (Parameter 1, Grid Midpoint)",
       xlab = "LRT Statistic", breaks = 30)
  abline(v = q95_chisquared, col = "red", lty = 2, lwd = 2)
  abline(v = q95_observed, col = "blue", lty = 2, lwd = 2)
  legend("topright", c("Chi-squared(1, 0.95)", "Observed 95% quantile"), col = c("red", "blue"), lty = 2)
  # Print the chi-squared quantile and the observed 95% quantile (also print the relative difference)
  cat(sprintf("Chi-squared(1, 0.95): %.4f\n", q95_chisquared))
  cat(sprintf("Observed 95%% quantile: %.4f\n", q95_observed))
  cat(sprintf("Relative difference: %.4f%%\n", 
              100 * (q95_observed - q95_chisquared) / q95_chisquared))
  
  # Plot LRT statistics across grid for first parameter
  allq95_observed = apply(result_bootstrap$bootstrap_stats[,1,], 2, quantile, probs = 0.95, na.rm = TRUE)
  plot(seq_len(dim(result_bootstrap$bootstrap_stats)[3]), allq95_observed, pch = 16, cex = 1.5, col = "blue",
       main = "Bootstrap LRT Statistics Across Grid (Parameter 1)",
       ylim = c(min(allq95_observed), 6),
       xlab = "Grid Point", ylab = "LRT Statistic")
  abline(h = q95_chisquared, col = "red", lty = 2, lwd = 2)
  legend("topright", c("Chi-squared(1, 0.95)", "Observed 95% quantile"), col = c("red", "blue"), lty = 2)
}
```
The difference between the thresholds will decrease with more bootstrap samples in this example.


## Results Comparison

```{r}
# Compare confidence intervals
cat("Fixed threshold (chi-squared) confidence intervals:\n")
print(result_fixed$confidence_intervals)

cat("\nBootstrap confidence intervals:\n")
print(result_bootstrap$confidence_intervals)

cat("\nConfidence interval widths:\n")
fixed_widths = result_fixed$confidence_intervals[,2] - result_fixed$confidence_intervals[,1]
bootstrap_widths = result_bootstrap$confidence_intervals[,2] - result_bootstrap$confidence_intervals[,1]

comparison = data.frame(
  Parameter = rownames(result_fixed$confidence_intervals),
  Fixed_Width = round(fixed_widths, 4),
  Bootstrap_Width = round(bootstrap_widths, 4),
  Ratio = round(bootstrap_widths / fixed_widths, 3)
)
print(comparison)
```

## Theoretical Comparison

For the normal distribution case, both methods should give similar results since the likelihood ratio statistics follow the asymptotic chi-squared distribution quite well:

```{r}
# Compare with theoretical confidence intervals for normal distribution
# Theoretical SE for mu: sigma/sqrt(n)
# Theoretical SE for sigma: sigma/sqrt(2*n) (approximately)
theoretical_se_mu = ml_params[2] / sqrt(n_data)
theoretical_se_sigma = ml_params[2] / sqrt(2*n_data)

# 95% theoretical CIs (using normal approximation)
z_025 = qnorm(0.975)
theoretical_ci_mu = ml_params[1] + c(-1, 1) * z_025 * theoretical_se_mu
theoretical_ci_sigma = ml_params[2] + c(-1, 1) * z_025 * theoretical_se_sigma

cat("Theoretical confidence intervals (normal approximation):\n")
cat(sprintf("mu:    [%.4f, %.4f]\n", theoretical_ci_mu[1], theoretical_ci_mu[2]))
cat(sprintf("sigma: [%.4f, %.4f]\n", theoretical_ci_sigma[1], theoretical_ci_sigma[2]))
```

## Summary

The bootstrap method provides:

- Data-driven thresholds that don't rely on asymptotic assumptions
- More robust confidence intervals for complex likelihood surfaces
- Similar results to chi-squared method when asymptotic conditions are met

For this normal distribution example, both methods give very similar results, validating the bootstrap implementation while demonstrating its use for more complex scenarios where asymptotic assumptions may not hold.
