---
title: "Profile Likelihood for a Gaussian Mixture Model"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Profile Likelihood for a Gaussian Mixture Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction to Profile Likelihood

Profile likelihood is a fundamental method for uncertainty quantification in statistical inference. Given a parameter vector $\theta = (\theta_1, \theta_2, \ldots, \theta_p)$ and a likelihood function $L(\theta)$, the profile likelihood for parameter $\theta_i$ is obtained by:

1. **Fixing** the parameter of interest $\theta_i$ at a specific value
2. **Optimizing** the likelihood over all other parameters $\theta_{-i} = \{\theta_1, \ldots, \theta_{i-1}, \theta_{i+1}, \ldots, \theta_p\}$
3. **Recording** the maximized likelihood value as a function of $\theta_i$

Mathematically, the profile likelihood for parameter $\theta_i$ is:

$$L_p(\theta_i) = \max_{\theta_{-i}} L(\theta_i, \theta_{-i})$$

Confidence intervals are then constructed using the likelihood ratio test statistic:

$$\text{LR}(\theta_i) = 2[\log L(\hat{\theta}) - \log L_p(\theta_i)]$$

where $\hat{\theta}$ is the maximum likelihood estimate. Under regularity conditions, $\text{LR}(\theta_i)$ follows a $\chi^2_1$ distribution, so 95% confidence intervals correspond to the region where $\text{LR}(\theta_i) \leq 3.84$.

## Test Case: Gaussian Mixture Model

This vignette demonstrates profile likelihood computation for a **two-component Gaussian mixture model** with correlated parameters. The model is defined by the negative log-likelihood:

$$-\log L(x; \gamma) = -\log\left[\exp\left(-\gamma (x - m_1)^T \Sigma_1^{-1} (x - m_1)\right) + \exp\left(-\gamma (x - m_2)^T \Sigma_2^{-1} (x - m_2)\right)\right]$$

where:

- $x \in \mathbb{R}^d$ is the parameter vector of interest (in our case, $d = 5$)
- $\gamma > 0$ is the concentration parameter (higher values create sharper peaks)
- $m_1 = (1, 1, 0, 0, 0)^T$ and $m_2 = (0, 0, 0, 0, 0)^T$ are the fixed mixture centers
- $\Sigma_1, \Sigma_2$ are covariance matrices that introduce correlations between parameters

We define this negative log-likelihood function in R as follows:
```{r}
# Problem parameters
gamma = 8   # concentration parameter
d = 5       # dimensionality

# Define covariance matrices to introduce correlations
Sigma1 = diag(d)
Sigma1[1, 2] = Sigma1[2, 1] = 0.5

# Pre-compute Cholesky decomposition of inverse covariance matrices
chol_sigma1_inv = chol(solve(Sigma1))
chol_sigma2_inv = chol_sigma1_inv  # Use the same covariance for both components

# Define the Gaussian mixture negative log-likelihood
mlogf = function(x, gamma, chol_sigma1_inv, chol_sigma2_inv){
  m1 = m2 = rep(0, length(x))
  m1[1] = 1
  m1[2] = 1
  diff1 = x - m1
  diff2 = x - m2
  - log(exp(- gamma * crossprod(chol_sigma1_inv %*% diff1)) + 
        exp(- gamma * crossprod(chol_sigma2_inv %*% diff2)))
}

```

For this example, let us define the true parameters as known, and some dummy optimized parameters as if we obtained them from an optimization routine.

```{r}
true_params = c(1, 1, rep(0, d-2))
# Define optimized parameters, slightly away from the true values
set.seed(23)
optimized_params = true_params + rnorm(length(true_params), sd = 0.05)
names(true_params) = names(optimized_params) = paste0("param_", 1:d)
```

Now we can perform profile likelihood computation using the `computeLikelihoodProfiles` function. See the documentation of the function for options.

```{r}
library(likelyprofiler)

# Bounds for the parameters
bounds = list(lower = rep(-2, d), upper = rep(2, d))

# Compute profile likelihood
result = computeLikelihoodProfiles(
  params_current = optimized_params,
  negLogLikelihood = mlogf,
  bounds = bounds,
  gamma = gamma,
  chol_sigma1_inv = chol_sigma1_inv,
  chol_sigma2_inv = chol_sigma2_inv
)

```

Print a summary of the results and confidence intervals:
```{r}
# View results
print(result$summary)
print(result$confidence_intervals)
```

Finally, we can visualize the profile likelihood using the `plotLikelihoodProfiles` function. This will show how the likelihood varies with each parameter, along with confidence intervals. This can help identify issues like parameter identifiability, or even issues with the optimization routine itself.

```{r}
# Plot the profile likelihood
plotLikelihoodProfiles(result,
                       true_values = true_params)
```

## Parallel Computation

The `computeLikelihoodProfiles` function supports parallel computation to speed up the profile likelihood calculation.

```{r, eval=FALSE}
# Create a cluster with 2 cores
cluster = parallel::makeCluster(2)

# Compute profile likelihood in parallel
result_parallel = computeLikelihoodProfiles(
  params_current = optimized_params,
  negLogLikelihood = mlogf,
  bounds = bounds,
  gamma = gamma,
  chol_sigma1_inv = chol_sigma1_inv,
  chol_sigma2_inv = chol_sigma2_inv,
  cluster = cluster
)

# Stop the cluster after computation
parallel::stopCluster(cluster)

```
