---
title: "Likelihood Profiling in Parallel"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Likelihood Profiling in Parallel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette demonstrates the parallel computation capabilities of the `likelyprofiler` package. For background on profile likelihood theory and the mathematical formulation, please refer to the "Profile Likelihood for a Gaussian Mixture Model" vignette. Here we focus on three parallelization strategies:

1. **Parameter-level parallelization**: Computing profiles for different parameters in parallel
2. **DEoptim parallelization**: Using DEoptim's built-in parallel capabilities  
3. **Nested parallelization**: Combining both approaches for maximum performance

We'll use the same correlated Gaussian mixture model to demonstrate performance gains from each approach.

## Problem Setup

First, let's set up the same correlated Gaussian mixture model from the basic vignette:

```{r problem_setup}
library(likelyprofiler)

# Problem parameters
gamma = 8   # concentration parameter
d = 5       # dimensionality

# Define covariance matrices to introduce correlations
Sigma1 = diag(d)
Sigma1[1, 2] = Sigma1[2, 1] = 0.5

# Pre-compute Cholesky decomposition of inverse covariance matrices
chol_sigma1_inv = chol(solve(Sigma1))
chol_sigma2_inv = chol_sigma1_inv  # Use the same covariance for both components

# Define the Gaussian mixture negative log-likelihood
mlogf = function(x, gamma, chol_sigma1_inv, chol_sigma2_inv){
  m1 = m2 = rep(0, length(x))
  m1[1] = 1
  m1[2] = 1
  diff1 = x - m1
  diff2 = x - m2
  - log(exp(- gamma * crossprod(chol_sigma1_inv %*% diff1)) + 
        exp(- gamma * crossprod(chol_sigma2_inv %*% diff2)))
}

# True and optimized parameters
true_params = c(1, 1, rep(0, d-2))
set.seed(23)
optimized_params = true_params + rnorm(length(true_params), sd = 0.05)
names(true_params) = names(optimized_params) = paste0("param_", 1:d)

# Bounds for the parameters
bounds = list(lower = rep(-2, d), upper = rep(2, d))

# Use smaller grid for faster demonstration
profile_options = list(
  grid_method = "adaptive",
  grid_points = 50,  # Reduced for faster execution
  max_grid_range_multiplier = 1.5
)
```

## Sequential Computation (Baseline)

First, let's establish a baseline with sequential computation:

```{r sequential}
# Sequential computation for timing comparison
system.time({
  result_sequential = computeLikelihoodProfiles(
    params_current = optimized_params,
    negLogLikelihood = mlogf,
    bounds = bounds,
    profile_options = profile_options,
    optimizer = "optim",
    verbose = FALSE,
    gamma = gamma,
    chol_sigma1_inv = chol_sigma1_inv,
    chol_sigma2_inv = chol_sigma2_inv
  )
})

print(result_sequential$summary)
```

## Parameter-Level Parallelization

The most straightforward parallelization approach is to compute profiles for different parameters simultaneously:

```{r param_parallel}
# Create a cluster for parameter-level parallelization
cluster = parallel::makeCluster(2)

# Compute profiles in parallel across parameters
system.time({
  result_param_parallel = computeLikelihoodProfiles(
    params_current = optimized_params,
    negLogLikelihood = mlogf,
    bounds = bounds,
    profile_options = profile_options,
    optimizer = "optim",
    cluster = cluster,
    verbose = FALSE,
    gamma = gamma,
    chol_sigma1_inv = chol_sigma1_inv,
    chol_sigma2_inv = chol_sigma2_inv
  )
})

print(result_param_parallel$summary)

# Stop the cluster
parallel::stopCluster(cluster)
```

## DEoptim with Parallel Function Evaluation

For more complex optimization problems, DEoptim can provide better results. DEoptim also supports parallel evaluation of the objective function:

```{r deoptim_parallel}
# Create cluster for DEoptim parallelization
cluster_deoptim = parallel::makeCluster(2)

# Use DEoptim with parallel function evaluation
system.time({
  result_deoptim_parallel = computeLikelihoodProfiles(
    params_current = optimized_params,
    negLogLikelihood = mlogf,
    bounds = bounds,
    profile_options = profile_options,
    optimizer = "deoptim",
    optim_options = list(
      itermax = 50,      # Reduced for faster demo
      NP = 50,           # Population size
      trace = FALSE,
      cluster = cluster_deoptim  # DEoptim's parallel evaluation
    ),
    verbose = FALSE,
    gamma = gamma,
    chol_sigma1_inv = chol_sigma1_inv,
    chol_sigma2_inv = chol_sigma2_inv
  )
})

print(result_deoptim_parallel$summary)

# Stop the cluster
parallel::stopCluster(cluster_deoptim)
```

## Nested Parallelization

For maximum performance on multi-core systems, we can combine both approaches using nested parallelization:

```{r nested_parallel, eval=FALSE}
# Create outer cluster for parameter-level parallelization
outer_cluster = parallel::makeCluster(2)

# Define nested cluster creation expression
# This will be evaluated on each worker to create inner clusters
nested_cluster_expr = quote({
  nested_cluster = parallel::makeCluster(2)
})

# Nested parallelization: parameters in parallel + DEoptim in parallel
system.time({
  result_nested_parallel = computeLikelihoodProfiles(
    params_current = optimized_params,
    negLogLikelihood = mlogf,
    bounds = bounds,
    profile_options = profile_options,
    optimizer = "deoptim",
    optim_options = list(
      itermax = 50,
      NP = 20,
      trace = FALSE
    ),
    cluster = outer_cluster,
    nested_cluster_call = nested_cluster_expr,
    verbose = FALSE,
    gamma = gamma,
    chol_sigma1_inv = chol_sigma1_inv,
    chol_sigma2_inv = chol_sigma2_inv
  )
})

print(result_nested_parallel$summary)

# Stop the outer cluster
parallel::stopCluster(outer_cluster)
```

## Conclusion

The `likelyprofiler` package offers flexible parallelization options to significantly speed up profile likelihood computation:

- **Parameter-level parallelization** is the easiest to implement and provides good speedup for most problems
- **DEoptim parallelization** is beneficial when likelihood evaluation is expensive  
- **Nested parallelization** can provide maximum performance but requires careful resource management

Choose the approach that best matches your problem characteristics and available computational resources. Always monitor resource usage to avoid assigning more cores than available, which can actually slow down computation.
